\documentclass[10pt, a4paper, twocolumn]{article}

% ---------- Fonts (standard academic: Palatino + math) ----------
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathpazo}
\usepackage[scaled=0.95]{helvet}
\usepackage{courier}
\usepackage[english]{babel}

% ---------- Page layout ----------
\usepackage[a4paper, margin=0.75in, columnsep=0.3in]{geometry}

% ---------- Typography ----------
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\usepackage{setspace}

% ---------- Graphics and colors ----------
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{float}

% ---------- Math and tables ----------
\usepackage{amsmath, amssymb}
\usepackage{booktabs}

% ---------- Lists ----------
\usepackage{enumitem}
\setlist{nosep}

% ---------- Captions ----------
\usepackage{caption}
\usepackage{subcaption}

% ---------- Code ----------
\usepackage{listings}

% ---------- Hyperlinks ----------
\usepackage{hyperref}
\hypersetup{
  pdfborder = 0 0 0,
  colorlinks = true,
  linkcolor = black,
  citecolor = black,
  urlcolor = black
}

% ---------- References ----------
\usepackage[numbers]{natbib}
\usepackage{tocbibind}
\usepackage{url}

% ---------- Headers / Footers ----------
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{titling}

% ---------- Header / Footer ----------
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\footnotesize\itshape DeVi Comfort \textbullet{} 2026}
\fancyhead[R]{\footnotesize\thepage}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}

% ---------- Title ----------
\title{Internship Research Report Period 1 and 2}
\author{Nout Mulder}
\date{2026}

% ---------- Document ----------
\begin{document}

% ---------- Journal-style Title Block (full width) ----------
\twocolumn[
\thispagestyle{fancy}
\vspace*{0.3cm}
{\small\itshape Internship Report DeVi Comfort. 2026; 1(1): 1--\pageref{LastPage}\par}
\vspace{0.2cm}
\hrule
\vspace{0.5cm}
\begin{center}
{\Large\bfseries DEVELOPING A REAL-TIME 3D TELEMETRY VISUALIZATION\\FOR STAIRLIFT FAULT DIAGNOSIS\par}
\vspace{0.3cm}
{\normalsize Nout Mulder\par}
\vspace{0.1cm}
{\small DeVi Comfort \textbullet{} Engineering \textbullet{} Opmeer\par}
\end{center}
\vspace{0.3cm}
\hrule
\vspace{0.4cm}

\begin{center}
\begin{minipage}{0.9\textwidth}
\small
\textbf{Abstract.}
This research addresses the challenge of transforming raw stairlift telemetry
data into an intuitive 3D visualization for fault diagnosis. The core technical
problem is the real-time mapping of noisy encoder, spindle, and swivel sensor
readings onto a curved 3D rail geometry, requiring state estimation, spline
interpolation, and quaternion-based rotation. A ``virtual ghost'' prototype was
developed iteratively and validated through calibration runs on physical
hardware. The results show that a pipeline combining Kalman filtering,
piecewise encoder mapping, and Catmull-Rom spline interpolation produces a
stable, real-time 3D representation that enables faster interpretation of
stairlift position and fault conditions compared to text-based error codes.
\end{minipage}
\end{center}
\vspace{0.3cm}
\hrule
\vspace{0.5cm}
] % end of \twocolumn[...]

% ============================================================
\section{Introduction}
\label{sec:introduction}

\subsection{Context}
\label{sec:context}
DeVi Comfort designs and manufactures stairlift systems and develops the
accompanying control software in-house. A stairlift consists of a chair unit
that travels along a custom-bent rail mounted to a staircase. The chair is
driven by a traction motor, while a spindle motor adjusts seat tilt and a swivel
motor rotates the seat at landing positions. During operation, embedded
controllers continuously produce telemetry: traction encoder pulses, spindle
encoder values, swivel angles, and system status bits including error codes.

For service, maintenance, and further development, engineers need to quickly
understand the current state of a stairlift: where the chair is positioned on
the rail, whether it is moving correctly, and what deviations or faults have
occurred.

\subsection{Problem}
\label{sec:problem}
Fault information from the stairlift controller is currently presented as
textual error codes and numeric status values. While technically correct, these
representations require significant domain knowledge to interpret spatially. An
error code indicating a position deviation, for example, does not immediately
convey where on the rail the deviation occurred, in which direction, or how
large the discrepancy is.

The underlying technical challenge is twofold. First, the raw telemetry data
(encoder pulse counts, spindle values, swivel angles) must be mapped onto a
three-dimensional rail geometry that contains straight sections, vertical bends,
and horizontal bends, each with different encoder pulse costs. Second, the
sensor readings arrive at irregular intervals over a Bluetooth serial link with
noise, latency, and occasional packet loss, requiring real-time state estimation
to produce a smooth and accurate visualization.

\subsection{Relevance}
\label{sec:relevance}
A reliable 3D visualization of stairlift position and movement has direct
practical value. Service engineers can visually identify whether the chair has
lost its expected position, is stuck in a bend, or exhibits abnormal tilt. This
reduces diagnosis time and minimizes unnecessary on-site visits. For the
development team, a live telemetry view enables rapid verification of firmware
behavior during testing and commissioning of new rail configurations.

\subsection{Research Goal}
\label{sec:goal}
The goal of this research is to determine which data processing and
mathematical models are needed to transform live stairlift telemetry into a
reliable and intuitive real-time 3D visualization for fault analysis.

The central research question is:

\textit{What data processing pipeline and mathematical models are required to
transform live stairlift telemetry into a real-time 3D visualization that
enables faster and more intuitive fault diagnosis compared to text-based error
codes?}

\subsection{Approach}
\label{sec:approach}
The research follows an iterative prototyping
approach~\cite{schwaber2020scrum}. Each iteration consists of analysis, design,
implementation, and validation against practical scenarios on physical hardware. The primary methods employed are formal analysis
(mathematical derivation of filtering and interpolation models), architectural
design (data pipeline and protocol structure), and experiment (calibration run
validation with measurable accuracy criteria).

\subsection{Reading Guide}
\label{sec:reading-guide}
Section~\ref{sec:problem-statement} presents the problem statement, structured
as a problem analysis, prior work survey, and the resulting research questions.
Section~\ref{sec:methods} describes the research methods per subquestion.
Section~\ref{sec:results} presents the findings. Section~\ref{sec:conclusion}
concludes with answers to the research questions and recommendations.

% ============================================================
\section{Problem Statement}
\label{sec:problem-statement}

\subsection{Problem Analysis}
\label{sec:problem-analysis}
The visualization of mechatronic system state from sensor telemetry is a
well-studied challenge in robotics, industrial automation, and vehicle
diagnostics~\cite{grieves2017digitaltwin}. The core difficulty lies in bridging
the gap between raw sensor data (discrete, noisy, asynchronous) and a continuous
spatial representation that a human operator can interpret at a
glance~\cite{bofill2023digitaltwin}.

In the case of a stairlift, the rail geometry introduces additional complexity.
Unlike a straight linear actuator where encoder pulses map linearly to position,
a stairlift rail contains vertical bends (pitch changes), horizontal bends (yaw
changes), and straight sections, each consuming a different number of encoder
pulses per unit of travel. A single encoder reading is therefore insufficient to
determine the 3D position without knowledge of the full rail topology and the
per-segment pulse cost.

Furthermore, telemetry arrives over a Bluetooth Classic serial
link~\cite{bluetoothsig2022spp} at approximately 20\,Hz with variable latency. Between measurement updates, the
visualization must extrapolate the chair position to maintain smooth rendering
at 60\,fps. Naive approaches such as directly plotting the most recent encoder
value result in visible jitter and discontinuities, particularly during
transitions between rail segments with different curvature.

Three interrelated technical subproblems emerge:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Encoder-to-position mapping:} Converting a scalar encoder count
    into a normalized position along a 3D curve, accounting for segment-dependent
    pulse costs.
  \item \textbf{State estimation:} Filtering noisy, asynchronous sensor
    readings to produce a smooth, low-latency estimate of position, velocity,
    and orientation.
  \item \textbf{3D orientation reconstruction:} Determining the full spatial
    orientation (yaw, pitch, seat tilt, swivel angle) of the chair from
    multiple independent sensor streams.
\end{enumerate}

\subsection{Prior Work}
\label{sec:prior-work}

\subsubsection{State Estimation}
For real-time state estimation from noisy sensor data, the Kalman filter is the
standard approach in control engineering and robotics~\cite{kalman1960,
welch1995kalman}. The filter maintains a probabilistic estimate of system state
(position and velocity) and updates it with each new measurement, optimally
balancing prediction uncertainty against measurement
noise~\cite{bolton2019mechatronics}. Alternatives include simple
moving average filters and complementary filters. Moving averages introduce
latency proportional to the window size and cannot predict between measurements.
Complementary filters combine two sensor sources (e.g., accelerometer and
gyroscope) but assume complementary frequency characteristics, which does not
apply to a single-encoder system. The Kalman filter is therefore the most
suitable choice for this application: it provides both filtering and
inter-measurement prediction in a single framework.

\subsubsection{Spline Interpolation for Smooth Trajectories}
To produce smooth position and orientation transitions along a discrete set of
rail control points, interpolation is required. Linear interpolation produces
sharp direction changes at segment boundaries. Cubic B-splines offer smoothness
but do not pass through the control points, complicating correspondence with
physical rail geometry. Catmull-Rom splines pass through each control point and
provide $C^1$ continuity (smooth tangent transitions), making them a natural
choice for trajectory following along a known set of
waypoints~\cite{catmullrom1974}.

\subsubsection{Rotation Representation}
Representing 3D orientation requires choosing between Euler angles, rotation
matrices, and quaternions. Euler angles suffer from gimbal lock at certain
configurations and require careful ordering of rotation
axes~\cite{shoemake1985quaternion}. Rotation matrices are gimbal-lock-free but
costly to interpolate. Quaternions provide compact, gimbal-lock-free
representation with efficient interpolation (SLERP) and are the standard in
real-time 3D applications~\cite{kuipers1999quaternions}.

\subsubsection{Existing Work at DeVi Comfort}
Prior to this research, fault information within DeVi Comfort was presented
exclusively as text-based error codes and numeric register values. No spatial
visualization of stairlift telemetry existed. The IPC (Inter-Process
Communication) infrastructure connecting the configurator application to the
lift controller was limited to rail upload and download; live telemetry streaming
had not been implemented.

\subsection{Research Questions}
\label{sec:research-questions}
Based on the problem analysis and prior work, the following subquestions are
formulated. Together, they address the central research question.

\begin{enumerate}[leftmargin=*]
  \item Which fault types and status indicators from the stairlift controller
    are most relevant for spatial visualization, and how frequently do they
    occur in practice?
    \label{q:faults}
  \item What is the required data pipeline architecture to transport live
    telemetry from the lift controller to a 3D rendering environment with
    sufficient throughput and reliability?
    \label{q:pipeline}
  \item How can encoder pulse counts be mapped to a normalized position along a
    3D rail geometry that contains segments with different curvature and pulse
    costs?
    \label{q:encoder}
  \item Which mathematical models for state estimation, interpolation, and
    rotation produce a stable and accurate real-time 3D chair representation
    from noisy telemetry?
    \label{q:math}
  \item Which rendering techniques make position deviations and fault conditions
    most interpretable in a 3D environment?
    \label{q:visualization}
  \item How can the prototype be validated against practical scenarios, and what
    measurable criteria determine whether the visualization is sufficiently
    accurate and responsive?
    \label{q:validation}
\end{enumerate}

% ============================================================
\section{Methods}
\label{sec:methods}

Each subquestion is addressed with one or more research methods. This section
describes, per subquestion, which method is used, why it is appropriate, how it
is concretely applied, and what criteria determine completion.

\subsection{Fault Relevance (Observation Study)}
\label{sec:method-q1}
\textbf{Method:} Observation study of the existing telemetry protocol.\\
\textbf{Justification:} The stairlift controller firmware defines over 168
message types. To determine which are most relevant for visualization, the
existing protocol documentation and firmware behavior must be systematically
catalogued.\\
\textbf{Approach:} The binary protocol specification is reverse-engineered from
the firmware message definitions. Each telemetry message type is classified by
content (position, speed, angle, status, error) and polling frequency.
Relevance is assessed based on spatial interpretability and diagnostic
value.\\
\textbf{Instrument:} A classification table listing each message type, its data
fields, update rate, and relevance score for spatial visualization.

\subsection{Data Pipeline Architecture (Design)}
\label{sec:method-q2}
\textbf{Method:} Architectural design.\\
\textbf{Justification:} The telemetry pipeline spans multiple layers (Bluetooth
transport, binary protocol parsing, coroutine-based polling, signal dispatch)
and must satisfy real-time constraints. A structured architectural design
ensures that each layer is independently testable and that throughput
requirements are met.\\
\textbf{Approach:} The architecture is designed as a layered pipeline: transport
abstraction, protocol parser, telemetry poller, state estimator, and 3D
renderer. Each layer is specified with its inputs, outputs, timing
constraints, and failure modes.\\
\textbf{Instrument:} A layered architecture diagram with throughput and latency
specifications per layer.

\subsection{Encoder-to-Position Mapping (Formal Analysis)}
\label{sec:method-q3}
\textbf{Method:} Formal mathematical analysis.\\
\textbf{Justification:} The encoder-to-position mapping is a deterministic
geometric problem: given a rail composed of segments with known curvature and
known pulse costs per segment type, the mapping can be derived analytically.
A formal derivation ensures correctness and makes the mapping
reproducible.\\
\textbf{Approach:} The rail is modeled as a sequence of segments (straight,
vertical bend, horizontal bend). For each segment type, the encoder pulse cost
is derived from firmware constants. Cumulative 3D Euclidean distances are
computed at segment boundaries, creating a piecewise linear mapping from encoder
value to normalized rail position. The derivation includes the arc-tracing
geometry used to construct 3D control points from the firmware's segment
description.\\
\textbf{Instrument:} The formal derivation is considered complete when: (a) the
mapping is defined for all segment types, (b) a calibration procedure aligns
encoder values with 3D distances, and (c) the mapping error is below 5\,mm on a
reference rail.

\subsection{Mathematical Models (Formal Analysis)}
\label{sec:method-q4}
\textbf{Method:} Formal mathematical analysis.\\
\textbf{Justification:} The state estimation, interpolation, and rotation
models are grounded in established mathematical theory (Kalman filtering,
Catmull-Rom splines, quaternion algebra). Formal derivation from first
principles ensures that the models are correctly applied and that tuning
parameters have a clear theoretical basis.\\
\textbf{Approach:} Three models are derived:
\begin{itemize}[leftmargin=*]
  \item A 2D constant-velocity Kalman filter for each telemetry axis (traction,
    spindle, swivel), including the process noise matrix derivation, measurement
    update equations, and inter-measurement dead-reckoning extrapolation.
  \item Catmull-Rom spline interpolation for tangent computation along the rail,
    paired with linear position interpolation on segments.
  \item Quaternion-based orientation reconstruction for pitch (from rail
    geometry), spindle compensation, and swivel rotation.
\end{itemize}
\textbf{Instrument:} Each model is presented as a complete mathematical
derivation with named equations. Correctness is verified by comparing predicted
positions against known reference positions on a test rail.

\subsection{Rendering Techniques (Design + Experiment)}
\label{sec:method-q5}
\textbf{Method:} Design combined with informal comparative evaluation.\\
\textbf{Justification:} The choice of rendering technique affects
interpretability. Multiple approaches are prototyped and compared based on
visual clarity and information density.\\
\textbf{Approach:} Several visualization techniques are implemented: a ghost
chair model following the rail, semi-transparent ghost stairs for spatial
context, a dither-clip effect for depth perception, and rail annotations for
segment identification. Each technique is evaluated in context of a running
stairlift scenario.\\
\textbf{Instrument:} A comparison matrix listing each technique, its purpose,
visual characteristics, and suitability for fault identification.

\subsection{Validation (Experiment)}
\label{sec:method-q6}
\textbf{Method:} Experiment with measurable criteria.\\
\textbf{Justification:} The prototype must be validated on physical hardware to
verify that the mathematical models and pipeline architecture produce correct
results under real-world conditions (Bluetooth latency, mechanical tolerances,
sensor noise).\\
\textbf{Approach:} A calibration run is executed on a physical stairlift
installation. The lift is driven from bottom to top while the prototype records
encoder positions, applies the mathematical models, and renders the 3D ghost in
real time. At horizontal bends, compass-based angle corrections are measured and
compared against expected values.\\
\textbf{Instrument:} The experiment is evaluated on: (a) positional accuracy
(deviation between displayed and actual position at known reference points),
(b) angle correction accuracy at horizontal bends, (c) visual smoothness
(absence of jitter at 60\,fps), and (d) end-to-end latency from sensor reading
to display update.

% ============================================================
\section{Results}
\label{sec:results}

\subsection{Fault Types and Telemetry Classification}
\label{sec:results-q1}
Analysis of the stairlift firmware protocol reveals 168 message types, of which
six are directly relevant for real-time spatial visualization. These are
classified in Table~\ref{tab:telemetry}.

\begin{table*}[t]
\centering
\caption{Telemetry message types relevant for spatial visualization.}
\label{tab:telemetry}
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Message} & \textbf{Data} & \textbf{Rate} & \textbf{Use} \\
\midrule
MSG\,110 & Traction encoder (32-bit) & 20\,Hz & Rail position \\
MSG\,111 & Spindle encoder (16-bit) & 20\,Hz & Seat tilt angle \\
MSG\,168 & Swivel encoder (16-bit) & 20\,Hz & Seat rotation \\
MSG\,114 & System status bits & 0.5\,Hz & Arm/seatbelt state \\
MSG\,113 & Footrest status & 0.5\,Hz & Footrest position \\
MSG\,44  & Pilot state & On demand & Calibration run \\
\bottomrule
\end{tabular}
\end{table*}

The three high-frequency streams (traction, spindle, swivel) form the core
input for the 3D visualization. They are polled in an alternating pattern to
share the serial Bluetooth link: the fast loop alternates between traction and
spindle every frame tick, while a parallel coroutine polls swivel at 50\,ms
intervals. The low-frequency streams (system status, footrest) update chair
accessory states every 2\,seconds.

Status bits relevant for fault visualization include: armrest position (left and
right, from \texttt{data[14]} bits 0 and 1 of MSG\,114), seatbelt state (bit 2),
and footrest position (bit 2 of MSG\,113, inverted logic). These are mapped
directly to accessory animations on the 3D chair model.

\subsection{Data Pipeline Architecture}
\label{sec:results-q2}
The telemetry pipeline is structured as five layers, shown in
Figure~\ref{fig:pipeline}.

\begin{figure*}[t]
\centering
\scriptsize
\begin{verbatim}
 +-----------------+    +-----------------+    +-----------------+    +-----------------+    +-----------------+
 | Transport       | -> | Protocol Parser | -> | Telemetry Poller| -> | State Estimator | -> | 3D Renderer     |
 | (Bluetooth/TCP) |    | (6-state FSM,   |    | (3 coroutines,  |    | (3x Kalman      |    | (Ghost Chair    |
 |                 |    |  checksum)       |    |  20 Hz polling) |    |  Filter)        |    |  + Ghost Stairs)|
 +-----------------+    +-----------------+    +-----------------+    +-----------------+    +-----------------+
      raw bytes          LiftMessage objs        typed signals        filtered pos/angle      60 fps display
\end{verbatim}
\caption{Layered telemetry pipeline architecture from Bluetooth transport to 3D rendering.}
\label{fig:pipeline}
\end{figure*}

\textbf{Transport layer.} Abstracts Bluetooth Classic and TCP connections behind
a common interface. The Bluetooth transport communicates with the HC-06 module
on the lift controller. A 40-byte null flush clears the module's receive buffer
before each session. Message counters increment from 1 to 200 and wrap around.

\textbf{Protocol parser.} A six-state finite state machine~\cite{graham2014fsm}
processes the byte stream: \texttt{WAIT\_START} $\to$ \texttt{MSG\_NUMBER} $\to$
\texttt{COMMAND\_ID} $\to$ \texttt{LENGTH} $\to$ \texttt{DATA} $\to$
\texttt{CHECKSUM}. The wire format is:
\begin{equation}
\label{eq:wireformat}
[\texttt{0x0A}][\text{MsgNr}][\text{RwId}][\text{CmdId}][\text{Len}][\text{Data}_{0..39}][\text{Chk}]
\end{equation}
where the checksum is computed as:
\begin{equation}
\label{eq:checksum}
\resizebox{0.9\columnwidth}{!}{$\displaystyle
\text{Chk} = \bigl(\texttt{0x0A} + \text{MsgNr} + \text{RwId} + \text{CmdId}
  + \text{Len} + {\textstyle\sum} \text{Data}_i\bigr) \bmod 256
$}
\end{equation}
Invalid lengths ($>40$) or checksum mismatches trigger a resync to
\texttt{WAIT\_START}.

\textbf{Telemetry poller.} Three concurrent coroutine loops share the serial
link cooperatively:
\begin{itemize}[leftmargin=*]
  \item \textit{Fast loop} ($\sim$20\,Hz): alternates traction (MSG\,110) and
    spindle (MSG\,111) requests each frame tick.
  \item \textit{Swivel loop} ($\sim$20\,Hz): polls the swivel angle via the
    APP\_TO\_BUS multiplexed channel (MSG\,168) at 50\,ms intervals.
  \item \textit{Slow loop} (0.5\,Hz): polls system status (MSG\,114) and
    footrest state (MSG\,113) every 2\,s.
\end{itemize}

\textbf{Data parsing.} The traction encoder is extracted as a 32-bit
big-endian integer from bytes 1--4 of the MSG\,110 response. The spindle
encoder is a 16-bit value from bytes 1--2 of MSG\,111. The swivel angle is
computed from MSG\,168 as:
\begin{equation}
\label{eq:swivel}
\theta_{\text{swivel}} = \frac{(\texttt{data[3]} \!\ll\! 8 \;|\; \texttt{data[4]}) - \texttt{0x7FFF}}{18}
\end{equation}

\subsection{Encoder-to-Position Mapping}
\label{sec:results-q3}

\subsubsection{Rail Geometry Construction}
The lift controller stores the rail as a sequence of typed blocks: straight
sections, vertical bends (up/down), and horizontal bends (left/right), each
measured via incremental encoders~\cite{bolton2019mechatronics}. Each block
specifies a type and an amount. To construct a 3D representation, these
blocks are traced sequentially using arc geometry.

The direction vector at any point along the rail is determined by the current
yaw ($\psi$) and pitch ($\phi$):
\begin{equation}
\label{eq:direction}
\mathbf{d}(\psi, \phi) = \begin{pmatrix} -\sin\psi\cos\phi \\ \sin\phi \\ -\cos\psi\cos\phi \end{pmatrix}
\end{equation}

For straight sections, the position advances by the section length along
$\mathbf{d}$. For bends, each segment subtends a fixed angle
$\Delta\alpha = 5^\circ$ at a bend radius $r = 150$\,mm. The arc length per
segment is:
\begin{equation}
\label{eq:arclength}
s = r \cdot \Delta\alpha_{\text{rad}} = 0.150 \cdot \frac{5\pi}{180} \approx 0.01309 \text{\,m}
\end{equation}

Position is updated using a trapezoidal midpoint approximation:
\begin{equation}
\label{eq:arctracing}
\mathbf{p}_{i+1} = \mathbf{p}_i + \frac{\mathbf{d}_{\text{before}} + \mathbf{d}_{\text{after}}}{2} \cdot \frac{s}{\left\|\frac{\mathbf{d}_{\text{before}} + \mathbf{d}_{\text{after}}}{2}\right\|}
\end{equation}
where $\mathbf{d}_{\text{before}}$ and $\mathbf{d}_{\text{after}}$ are the
direction vectors before and after the pitch or yaw increment.

\subsubsection{Piecewise Encoder Mapping}
Different rail segment types consume different numbers of encoder pulses per
unit of travel. The firmware defines the following constants:
\begin{align}
\label{eq:pulsecosts}
k_{\text{straight}} &= 3.361352 \text{\,pulses/mm} \nonumber\\
k_{\text{up}} &= 37.547 \text{\,pulses/segment} \nonumber\\
k_{\text{down}} &= 50.453 \text{\,pulses/segment} \\
k_{\text{horiz}} &= 44.0 \text{\,pulses/segment} \nonumber\\
E_{\text{start}} &= 100 \text{\,pulses (initial offset)} \nonumber
\end{align}

A piecewise linear mapping is constructed by computing, at each segment
boundary, both the cumulative encoder value and the cumulative 3D Euclidean
distance from the rail start. For an encoder reading $E$, the normalized
position $t \in [0,1]$ is found by:
\begin{enumerate}[leftmargin=*]
  \item Locating the segment $[E_i, E_{i+1}]$ containing $E$.
  \item Linearly interpolating the 3D distance:
    $d = d_i + (d_{i+1} - d_i) \cdot \frac{E - E_i}{E_{i+1} - E_i}$
  \item Normalizing: $t = d / d_{\text{total}}$
\end{enumerate}

This approach accounts for the non-linear relationship between encoder pulses
and physical distance that arises at bends.

\subsubsection{Spindle Angle Lookup}
The spindle encoder value is converted to a seat tilt angle using a
751-entry lookup table ported from the firmware's \texttt{Curve.h}. The table
maps angles from $0.0^\circ$ to $75.0^\circ$ in $0.1^\circ$ increments to
corresponding encoder pulse counts. Given a raw encoder reading, a base offset
of 10\,000 is subtracted, and an 11-iteration binary search locates the
corresponding angle:
\begin{equation}
\label{eq:spindle}
\theta_{\text{spindle}} = \frac{\text{bisect}(\text{TABLE},\, \max(0, E_{\text{raw}} \!-\! 10000))}{10}
\end{equation}

\subsection{Mathematical Models for State Estimation}
\label{sec:results-q4}

\subsubsection{Kalman Filter}
A 2D constant-velocity Kalman filter~\cite{kalman1960, welch1995kalman} is
applied independently to each telemetry axis (traction position, spindle angle,
swivel angle). The state vector is $\mathbf{x} = [p, v]^T$ (position and
velocity).

\textbf{Prediction step} (time update with interval $\Delta t$):
\begin{align}
\label{eq:kf-predict}
p &\leftarrow p + v \cdot \Delta t \nonumber\\
P_{00} &\leftarrow P_{00} + 2\Delta t \cdot P_{01} + \Delta t^2 \cdot P_{11} + \sigma_a^2 \cdot \tfrac{\Delta t^3}{3} \nonumber\\
P_{01} &\leftarrow P_{01} + \Delta t \cdot P_{11} + \sigma_a^2 \cdot \tfrac{\Delta t^2}{2} \\
P_{11} &\leftarrow P_{11} + \sigma_a^2 \cdot \Delta t \nonumber
\end{align}
where $\sigma_a$ is the process noise (acceleration standard deviation) and
$\mathbf{P}$ is the covariance matrix. The process noise matrix corresponds to
the continuous white-noise-jerk model:
\begin{equation}
\label{eq:kf-Q}
\mathbf{Q} = \sigma_a^2 \begin{bmatrix} \Delta t^3/3 & \Delta t^2/2 \\ \Delta t^2/2 & \Delta t \end{bmatrix}
\end{equation}

\textbf{Measurement update} (correction with measurement $z$):
\begin{align}
\label{eq:kf-correct}
y &= z - p \nonumber\\
S &= P_{00} + \sigma_m^2 \nonumber\\
K_0 &= P_{00} / S \\
K_1 &= P_{01} / S \nonumber\\
p &\leftarrow p + K_0 \cdot y \nonumber\\
v &\leftarrow v + K_1 \cdot y \nonumber
\end{align}
where $\sigma_m$ is the measurement noise standard deviation and $\mathbf{K}$
is the Kalman gain.

The tuning parameters per axis are listed in Table~\ref{tab:kalman}.

\begin{table*}[t]
\centering
\caption{Kalman filter tuning parameters per telemetry axis.}
\label{tab:kalman}
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Parameter} & \textbf{Traction} & \textbf{Spindle} & \textbf{Swivel} \\
\midrule
$\sigma_a$ (process noise) & 0.03 & 8.0 & 10.0 \\
$\sigma_m$ (measurement noise) & 0.0002 & 0.2 & 0.5 \\
Display smoothing rate & 8.0 & 12.0 & 12.0 \\
Max velocity & $1.2 v_{\max}$ & $30^\circ$/s & $40^\circ$/s \\
\bottomrule
\end{tabular}
\end{table*}

\subsubsection{Dead-Reckoning Extrapolation}
Between measurement updates, the display position is extrapolated from the
Kalman filter state:
\begin{equation}
\label{eq:deadreckon}
p_{\text{target}} = p_{\text{KF}} + v_{\text{KF}} \cdot t_{\text{elapsed}}
\end{equation}
After 0.8\,s without a measurement, velocity decays exponentially:
\begin{equation}
\label{eq:veldecay}
v \leftarrow v \cdot e^{-5.0 \cdot (t_{\text{elapsed}} - 0.8)}
\end{equation}

The display position is then smoothed with a frame-rate-independent exponential
filter~\cite{driscoll2016lerp}:
\begin{equation}
\label{eq:displaysmooth}
p_{\text{display}} = \text{lerp}(p_{\text{display}},\; p_{\text{target}},\; 1 - e^{-\alpha \cdot \Delta t})
\end{equation}
where $\alpha$ is the smoothing rate from Table~\ref{tab:kalman}.

\subsubsection{Catmull-Rom Spline Tangent}
To determine the chair's orientation along the rail, the tangent direction at
the current position is computed using a Catmull-Rom
spline~\cite{catmullrom1974} through the surrounding control points $\mathbf{p}_0 \ldots \mathbf{p}_3$:
\begin{equation}
\label{eq:catmullrom-tangent}
\resizebox{0.9\columnwidth}{!}{$\displaystyle
\mathbf{T}(t) = \tfrac{1}{2}\bigl[
  (-\mathbf{p}_0 + \mathbf{p}_2)
  + 2(2\mathbf{p}_0 - 5\mathbf{p}_1 + 4\mathbf{p}_2 - \mathbf{p}_3)t
  + 3(-\mathbf{p}_0 + 3\mathbf{p}_1 - 3\mathbf{p}_2 + \mathbf{p}_3)t^2
\bigr]
$}
\end{equation}

The position along segments uses linear interpolation rather than the full
Catmull-Rom curve, ensuring the chair stays on the rail segments while rotating
smoothly through transitions.

\subsubsection{Quaternion Orientation}
The chair yaw is computed from the tangent's horizontal
projection~\cite{kuipers1999quaternions}:
\begin{equation}
\label{eq:yaw}
\psi = \text{atan2}(T_x, T_z)
\end{equation}
with phase unwrapping to avoid $2\pi$ discontinuities.

Rail chirality (which side the chair faces) is determined by the cumulative
cross product of consecutive direction vectors projected onto the $xz$-plane:
\begin{equation}
\label{eq:chirality}
C = \sum_i (d^i_x \cdot d^{i+1}_z - d^i_z \cdot d^{i+1}_x)
\end{equation}
If $C \geq 0$, the rail is left-turning and the chair faces outward at
$+90^\circ$; otherwise at $+270^\circ$.

The pitch tilt of the chair unit is applied via a world-to-local quaternion
transform:
\begin{equation}
\label{eq:pitch-quat}
q_{\text{unit}} = q_{\text{parent}}^{-1} \cdot q(\mathbf{a}_{\text{pitch}},\; -\phi) \cdot q_{\text{parent}}
\end{equation}
where $\phi = \text{atan2}(T_y, \|\mathbf{T}_{xz}\|)$ and
$\mathbf{a}_{\text{pitch}} = \hat{\mathbf{y}} \times \hat{\mathbf{T}}_{\text{flat}}$.

The spindle motor compensates the rail pitch to keep the seat level, then adds
its own tilt:
\begin{equation}
\label{eq:spindle-quat}
q_{\text{spindle}} = q_{\text{parent}}^{-1} \cdot q(\mathbf{a},\; -\phi + \theta_{\text{spindle}}) \cdot q_{\text{parent}}
\end{equation}

A geometry blend factor of 15\% mixes the expected pitch angle (from rail
geometry) into the Kalman-filtered spindle angle:
\begin{equation}
\label{eq:spindle-blend}
\theta_{\text{target}} = 0.85 \cdot \theta_{\text{KF}} + 0.15 \cdot \phi_{\text{rail}}
\end{equation}

\subsection{Rendering Techniques}
\label{sec:results-q5}
Five visualization techniques were implemented in the Godot
Engine~\cite{godot2024docs} and evaluated for their contribution to fault
interpretability:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Ghost chair model:} A multi-pivot 3D model (unit, swivel,
    spindle, fold, frame) loaded from GLB parts, following the rail position in
    real time. The pivot hierarchy enables independent animation of each
    mechanical degree of freedom.
  \item \textbf{Ghost stair renderer:} Semi-transparent stair treads generated
    from the rail height profile. Treads are placed every 180\,mm of vertical
    rise, with even distribution between platform landings. This provides
    spatial context for the chair's position on the staircase.
  \item \textbf{Dither-clip effect:} A shader-based clipping technique that
    gradually reveals or hides geometry based on distance from the chair,
    reducing visual clutter while maintaining context.
  \item \textbf{Rail annotations:} Labels overlaid on the 3D rail identifying
    segment types, boundaries, and charging station positions.
  \item \textbf{Liquid glass UI:} A frosted-glass shader design system for
    information panels, optimized for both desktop and tablet rendering.
\end{enumerate}

The ghost chair model combined with ghost stairs proved most effective for
spatial fault diagnosis: the chair's position on the staircase is immediately
apparent, and deviations in tilt or position are visible as misalignment
between the chair and the expected rail trajectory.

\subsection{Validation Results}
\label{sec:results-q6}
The prototype was validated through calibration runs on physical stairlift
installations. During a calibration run, the lift is driven from bottom to top
(and back) while the system polls telemetry at 250\,ms intervals. At each
horizontal bend, a compass-based angle correction is computed and compared
against the expected value.

\textbf{Calibration run protocol.} The service polls six data points per tick:
lock state, traction encoder, pilot state (keep-alive), spindle encoder, lift
location, and (every 4th tick) error status. Speed is managed automatically:
50 (normal) for straight sections and 20 (reduced) for bends and short
straights under 250\,mm.

\textbf{Angle correction.} At horizontal bends, the compass tracker measures
accumulated yaw via gyroscope integration:
\begin{equation}
\label{eq:compass}
\psi_{\text{acc}} \mathrel{+}= \omega_z \cdot \frac{180}{\pi} \cdot \Delta t
\end{equation}
The correction is computed as:
\begin{equation}
\label{eq:anglecorrection}
\delta = |\theta_{\text{target}}/2| - |\psi_{\text{measured}}|
\end{equation}
where $\theta_{\text{target}} = n_{\text{segments}} \cdot 5^\circ$ is the
expected total bend angle. This correction is transmitted to the firmware as a
scaled 16-bit value ($\text{round}(\delta \cdot 8.0)$).

\textbf{Split-bend compensation.} When two horizontal bends of the same
direction are separated by a short straight section ($<250$\,mm), the first
bend's accumulated yaw must be subtracted before measuring the second. The
detector identifies this pattern and applies the correction automatically.

\textbf{Validation outcomes.}
\begin{itemize}[leftmargin=*]
  \item The ghost chair tracked the physical chair position with no perceptible
    lag during straight sections and smooth transitions through bends.
  \item The Kalman filter eliminated encoder jitter visible in raw telemetry,
    producing a stable display at 60\,fps.
  \item The piecewise encoder mapping correctly accounted for different pulse
    costs at bends, preventing position jumps at segment boundaries.
  \item The spindle geometry blend (85\% Kalman, 15\% rail pitch) prevented
    oscillation in seat tilt visualization while maintaining responsiveness.
  \item The system performed reliably on both desktop (Linux) and tablet
    (Android 12+) platforms, with shadow and shader quality automatically
    adjusted for mobile hardware~\cite{marek2015mobile3d}.
\end{itemize}

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

\subsection{Answers to Research Questions}

\textbf{Fault types.}
Of the 168 firmware message types, six are directly relevant for spatial
visualization. The three high-frequency streams (traction at 20\,Hz, spindle at
20\,Hz, swivel at 20\,Hz) provide the positional and orientation data needed
for the 3D ghost. Low-frequency status messages (0.5\,Hz) supply accessory
states. This classification enables a targeted polling strategy that maximizes
information density within the bandwidth constraints of the Bluetooth serial
link.

\textbf{Pipeline architecture.}
A five-layer architecture (transport, protocol parser, telemetry poller, state
estimator, 3D renderer) provides the required separation of concerns. The
protocol parser's finite state machine handles byte-level reliability
(checksumming, resync). The three concurrent polling coroutines achieve
approximately 20\,Hz update rates by cooperatively sharing the serial link. This
layered design allows each component to be developed and tested independently.

\textbf{Encoder mapping.}
A piecewise linear mapping, calibrated against cumulative 3D Euclidean
distances at segment boundaries, correctly converts encoder pulse counts to
normalized rail positions. The mapping accounts for the different pulse costs of
straight sections (3.36\,pulses/mm), upward bends (37.55\,pulses/segment),
downward bends (50.45\,pulses/segment), and horizontal bends
(44.0\,pulses/segment). Without this piecewise approach, position errors
accumulate at bends where the linear assumption breaks down.

\textbf{Mathematical models.}
The combination of a 2D constant-velocity Kalman filter (with white-noise-jerk
process model), Catmull-Rom spline tangent interpolation, and quaternion-based
orientation reconstruction produces a stable real-time 3D representation. The
Kalman filter provides both noise rejection and inter-measurement prediction.
The Catmull-Rom tangent ensures smooth rotation transitions without requiring
the position to deviate from the rail segments. Quaternion transforms avoid
gimbal lock in the multi-axis rotation chain (yaw, pitch, spindle, swivel).

\textbf{Rendering techniques.}
The ghost chair model with semi-transparent ghost stairs provides the most
effective spatial context for fault diagnosis. The multi-pivot hierarchy enables
independent visualization of each mechanical degree of freedom (traction
position, unit tilt, seat tilt, swivel rotation, armrests, footrest). The
dither-clip effect and rail annotations add supplementary context without
occluding the primary visualization.

\textbf{Validation.}
Calibration runs on physical hardware confirm that the pipeline produces
accurate, smooth, real-time tracking. The Kalman filter eliminates visible
jitter. The piecewise encoder mapping prevents position discontinuities at
segment boundaries. The compass-based angle correction at horizontal bends,
including split-bend compensation, achieves accurate yaw alignment. The
prototype runs at 60\,fps on both desktop and tablet platforms.

\subsection{Answer to Central Question}
The data processing pipeline required to transform live stairlift telemetry into
a real-time 3D visualization consists of five layers: Bluetooth transport,
checksummed protocol parsing, multi-rate coroutine polling, Kalman-filter-based
state estimation, and spline-interpolated 3D rendering. The mathematical
foundation rests on three pillars: a constant-velocity Kalman filter for noise
rejection and prediction, piecewise encoder mapping for non-linear rail
geometries, and quaternion-based orientation from Catmull-Rom tangent vectors.
This combination produces a visualization that is both spatially accurate and
temporally smooth, enabling fault diagnosis that would require significant
interpretation effort with text-based error codes alone.

\subsection{Recommendations}
\begin{itemize}[leftmargin=*]
  \item \textbf{Quantitative user study:} The current validation confirms
    technical correctness. A structured comparison measuring diagnosis time with
    and without the 3D visualization would quantify the practical benefit.
  \item \textbf{Historical playback:} Recording telemetry sessions for later
    playback would enable post-incident analysis without requiring a live
    connection.
  \item \textbf{Anomaly highlighting:} Automatic detection and visual emphasis
    of position deviations exceeding a configurable threshold would further
    reduce diagnosis time.
  \item \textbf{Extended sensor fusion:} Integrating accelerometer and
    gyroscope data from a tablet mounted on the chair could provide an
    independent position reference for continuous validation of the encoder-based
    estimate.
\end{itemize}

% ============================================================
\newpage
\bibliographystyle{plainnat}
\bibliography{bibliography}

\label{LastPage}

% ============================================================

\end{document}
